---
title: "Exploratory Data Analysis"
author: "Justin Priest"
date: "September 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('Fish692-PrudhoeFish_dataimport.R')
```


## Intro
This document summarizes some of the exploratory data analysis of the Prudhoe Bay nearshore
fisheries data by Justin Priest for Franz Mueter's Data Wrangling Seminar, Fish 692  
<https://github.com/justinpriest/Fish692-DataWranglingSeminar>

## Site Locations
```{r locations, echo=FALSE, warning=FALSE, message=FALSE}
library(leaflet)

pru_loc <- data.frame(Station = as.character(c(214, 218, 220, 230)),
           Name = c("Niakuk", "West Beach", "West Dock", "Endicott"),
           Lat  = c(70.347, 70.358, 70.3826, 70.309),
           Long = c(-148.210, -148.473, -148.558, -147.884) )

places <-  data.frame(Place = c("Prudhoe Bay", "Sag River Delta"),
           Lat = c(70.345, 70.31),
           Long = c(-148.38, -148.1))

leaflet(pru_loc) %>% addTiles() %>% setView(-148.2, 70.35, zoom = 10) %>%
  addMarkers(popup = ~paste0(Station, " - ", Name)) %>%
  addLabelOnlyMarkers(data = places, label = ~Place,
    labelOptions = labelOptions(noHide = T, textOnly = TRUE, 
                                textsize = "16px", direction = "bottom"))

```
The two eastern sites are 230 and 214, which are within the zone of influence of the Sag River.  

## Catch Histograms for select species (Response Variables)

Note that most catches are very right skewed
```{r responseplots, echo=FALSE, warning=FALSE, message=FALSE}


###########################
### Explore Response Variables ####
ggplot(allcatch %>% filter(Species=="ARCS" & totcount>0), aes(x=(totcount))) + geom_histogram(bins = 30) + 
  ggtitle("Arctic Cisco catch histogram (excluding zeros)")
ggplot(allcatch %>% filter(Species=="ARCD" & totcount>0), aes(x=totcount)) + geom_histogram(bins = 30) + 
  ggtitle("Arctic Cod catch histogram (excluding zeros)")
ggplot(allcatch %>% filter(Species=="BDWF" & totcount>0), aes(x=totcount)) + geom_histogram(bins = 30) + 
  ggtitle("Broad Whitefish catch histogram (excluding zeros)")
ggplot(allcatch %>% filter(Species=="BDWF" & totcount>0), aes(x=log(totcount))) + geom_histogram(bins = 30) + 
  ggtitle("Broad Whitefish catch histogram, log transformed (excluding zeros)")
# note the slightly better distribution with the log transform
ggplot(allcatch %>% filter(Species=="LSCS" ), aes(x=(day.of.year), y=totcount)) + geom_point() +
  facet_wrap(~Year, scales = "free") + ggtitle("Least Cisco catch by day of year") + ylab("Count")
# summary: most catches are of zeros or very low catch abundance, very right skewed
```

******  
## Environmental Histograms (explanatory variables)
\newline

```{r explanplots1, echo=FALSE, warning=FALSE, message=FALSE}

###########################
### Explore Explanatory Variables ####

ggplot(watersalin %>% group_by(Year) %>% 
         summarise(annsal=mean(c(Salin_Top, Salin_Mid, Salin_Bot), na.rm = TRUE)),
       aes(x=Year, y=annsal)) + geom_line() + ylab("Mean annual salinity (ppt)")

ggplot(watertemps %>% group_by(Year) %>% 
        summarise(anntemp=mean(c(Temp_Top, Temp_Mid, Temp_Bot), na.rm = TRUE)),
      aes(x=Year, y=anntemp)) + geom_line() + ylab("Mean annual temperature (C)")

#show how salinity has been changing
plottext_salin <- data.frame(label = c(
    paste0("annual change: ", signif(coef(summary(lm(
      Salin_Mid ~ Year, data=watersalin %>% filter(Station == 214))))[2,1], 2)),
    paste0("annual change: ", signif(coef(summary(lm(
      Salin_Mid ~ Year, data=watersalin %>% filter(Station == 218))))[2,1], 2)),
    paste0("annual change: ", signif(coef(summary(lm(
      Salin_Mid ~ Year, data=watersalin %>% filter(Station == 220))))[2,1], 2)),
    paste0("annual change: ", signif(coef(summary(lm(
      Salin_Mid ~ Year, data=watersalin %>% filter(Station == 230))))[2,1], 2)) ),
  Station   = c(214, 218, 220, 230) )

ggplot(watersalin %>% filter(Station != 231), aes(as.factor(Year), Salin_Mid)) + 
  geom_boxplot() + scale_y_continuous(limits = c(0,35)) +
  facet_wrap(~Station, nrow=2) +
  geom_smooth(method="lm", aes(group=1)) +
  geom_text(data = plottext_salin, 
            mapping = aes(x=-Inf, y=-Inf, label = label),
            hjust = -0.2, vjust = -15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# salinity has significantly dropped for the eastern sites (230&214) 
# and increased at the western sites (220&218)
```

\newline
Note that salinity has dropped at the eastern sites (230&214) 
and increased at the western sites (220&218)
\newline

```{r explanplots2, echo=FALSE, warning=FALSE, message=FALSE}
#show how temps have been increasing:
plottext_temp <- data.frame(label = c(
  paste0("annual change: ", signif(coef(summary(lm(
    Temp_Mid ~ Year, data=watertemps %>% filter(Station == 214))))[2,1], 2)),
  paste0("annual change: ", signif(coef(summary(lm(
    Temp_Mid ~ Year, data=watertemps %>% filter(Station == 218))))[2,1], 2)),
  paste0("annual change: ", signif(coef(summary(lm(
    Temp_Mid ~ Year, data=watertemps %>% filter(Station == 220))))[2,1], 2)),
  paste0("annual change: ", signif(coef(summary(lm(
    Temp_Mid ~ Year, data=watertemps %>% filter(Station == 230))))[2,1], 2)) ),
  Station   = c(214, 218, 220, 230) )

ggplot(watertemps %>% filter(Station != 231), aes(as.factor(Year), Temp_Mid)) + 
  geom_boxplot() + scale_y_continuous(limits = c(0,16)) +
  facet_wrap(~Station, nrow=2) +
  geom_smooth(method="lm", aes(group=1)) +
  geom_text(data = plottext_temp, 
            mapping = aes(x=-Inf, y=-Inf, label = label),
            hjust = -0.2, vjust = -15) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
#all 4 sites have had a significant increase in temperatures
```

\newline
Note that temperature has significantly increased at all sites!
\newline

```{r explanplots3, echo=FALSE, warning=FALSE, message=FALSE}
#Wind 
ggplot(deadhorsewind %>% filter(month==7 | month==8), aes(x=dailymeandir)) + 
  geom_histogram(col="black", bins = 40) + coord_polar() + 
  ggtitle("polar histogram plot of wind directions") + xlab("sustained wind direction")

```
